Covariate analysis of UFL-Mayo-ISB RNASeq temporal cortex

### 2016-01-30 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq from cerebellum"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB cerebellum RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn5201012';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

@@

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA <- METADATA %>%
  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER")))
```
Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER, based on the eucledian distance from other samples and/or seperated on the first 2 PC's
```{r preprocess.data1}
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
ContCovariates <- c("RIN", "AgeAtDeath")

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Gene Filtering
Retain genes with atleast 50% of the samples have >= 1 gene counts
```{r gene.filtering, results='asis', echo=FALSE}
# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(EXPR, 
                                      MIN_GENE_CPM=1,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5,
                                      verbose=T)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),
                                     preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('Diagnosis')

@@

```{r adjusted.voom.normalisation, results='asis', echo=FALSE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)
writeLines('Degenerate variables are:')
writeLines(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols]

@@

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
                          design=DM1$design, 
                          weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate.residuals, include=FALSE, results = 'asis'}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                              VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Add "Diagnosis"" back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic_Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(summary(COVARIATES$Diagnosis))
colnames(tmp) = "No. of samples"
kable(tmp)
```
```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

@@

DE = ldply(DE, .id = 'Comparison')
tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
kable(tmp)
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum RNASeq', parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq from cerebellum"
author: "Thanneer Perumal"
date: '`r date()`'
output:
  html_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(plyr)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

@@

```{r synapse_parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB cerebellum RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download_data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn5201012';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

@@

### Data preprocessing
```{r preprocess_data, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA <- METADATA %>%
  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER")))
```
Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER, based on the eucledian distance from other samples and/or seperated on the first 2 PC's
```{r preprocess_data1}
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates_clustering}
FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
ContCovariates <- c("RIN", "AgeAtDeath")

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates_correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Gene Filtering
Retain genes with atleast 50% of the samples have >= 1 gene counts
```{r gene_filtering, results='asis', echo=FALSE}
# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(EXPR, 
                                      MIN_GENE_CPM=1,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5,
                                      verbose=T)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial_voom_normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse_normalise_data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted_covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),
                                     preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars_NULL_ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative_adjusted_voom_normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('Diagnosis')

@@

```{r adjusted_voom_normalisation, results='asis', echo=FALSE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)
writeLines('Degenerate variables are:')
writeLines(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols]

@@

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
                          design=DM1$design, 
                          weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate_residuals, include=FALSE, results = 'asis'}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                              VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars_ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Add "Diagnosis"" back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic_Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse_normalise_data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(summary(COVARIATES$Diagnosis))
colnames(tmp) = "No. of samples"
kable(tmp)
```
```{r get_gene_symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

@@

DE = ldply(DE, .id = 'Comparison')
tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
kable(tmp)
```
### Store files in synapse
```{r synapse_store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum RNASeq', parentId = parentId)
CODE <- synStore(CODE)

### 2016-01-30 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Minor modifications

knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_RNASeq.Rmd",
                                 parentId ="syn5570325",
                                 entityName = 'UFL_Mayo_ISB Temporal Cortex RNASeq')
```

@@

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB temporal cortex RNASeq data';

@@

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(COVARIATES), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
collectGarbage()
dev.off()
```

>>>>

knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_RNASeq.Rmd",
                                 parentId ='syn5612983',#"syn5570325",
                                 entityName = 'UFL_Mayo_ISB Temporal Cortex RNASeq')
```

@@

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(fpc)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5612983'#'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB temporal cortex RNASeq data';

@@

# Cluster expression
clust = cutreeStatic(sampleTrees, cutHeight = 100, minSize = 10)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       clust = clust)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(clust), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]

plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES)), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
collectGarbage()
dev.off()

writeLines(paste('Samples that deviate from others are', paste(sampleTrees$labels[clust==0], collapse = ',')))
```
